Investigating the Effect of Dataset Representivity for Image Colourisation

### 2019-10-13 Paper/paper.tex
Edited paper

Sufficient research exists for the implementation of bias-mitigating machine learning algorithms, however, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general. To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise greyscale images of a South African context. There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to 75 epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after 5 epochs. This did lead to the hues of the images being slightly more South African in essence, but which was largely unnoticed by the human eye. Consequently, the inclusion of local data in the dataset definitely minimises the bias of the dataset, but whether or not that bias-mitigation justifies the vast resources required to build the comprehensive dataset is determined on a case-by-case basis.

@@

The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being initially used, particularly a non-South African dataset being used to train a model to operate on a South African sample.

Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset. However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required. Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled nor does it require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.

@@

For simplicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space. This makes it convenient when separating the grey and colour channels due to the fact that the 'L' channel in LAB is simply the greyscale image. This greyscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image. These 'AB' channels make up the colour and simply signify the red-green and blue-yellow shift respectively. HSB would be another option when choosing a suitable colour space. In general, the choice of colour space makes no difference to the resulting image that our network produces.

@@

When colourising images using a deep learning approach, it is generally easiest to develop the model as a regression problem. Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland. This is the result of the multi-modality of pictures: there is no clear colour for any one image. This leads our model, using an aggregate loss function, to simply find an average colour, resulting in a bland image.

@@

For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand. We utilize a mean squared error loss which may average colour distribution but still produces images with a reasonable amount of colour. If we aimed to produce images which are more diverse in colour and saturated, regression would not suffice. Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the output and the representivity of local images in the data, which it does.

@@

Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al. \cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colourisation'\cite{zhang2016colorful} approached colourisation as a classification problem.

Our network topology consists of the first six layers of a ResNet-18, followed by three layers of deconvolutions. We find that this simple structure works well enough for our purposes and is relatively fast. Our model uses the Adam optimiser alongside PyTorch's autograd which makes backpropogating in our network trivial.

@@

In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 30\% inclusive, increasing in graduations of 10\%. In addition to this training dataset, a validation dataset (of 1000 images in total) is also prepared with the same corresponding percentage inclusion of South African images.

Each version of the dataset will be used to train the model independently of each other for a total of 75 epochs each, with validation being done at the conclusion of each epoch.

A set collection of South African and International images will be passed through each model respectively, with the subsequent losses being an indication of the bias present in the model due to the predominant locale present in the training images (we take the dataset version with 0\% South African images included as the base case of how the model performs on South African image evaluation when trained exclusively on International images).

@@

The collection of South African images\footnote{All images are freely available with a Creative Commons Licence or are in the public domain} was downloaded from the image-sharing service, Flickr. Only images which have been tagged with one of the following keywords were selected:

@@

A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 30\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split. The same number of images ($x$\%) was removed from the foreign dataset meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30\}$

@@

For each dataset, we trained our convolutional model from the beginning with the exact same hyperparameters. The following table illustrates the hyperparameters used:

@@

Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a higher inclusion rate to the above mentioned epoch. For this reason, we chose to limit our inclusion of South African data to a maximum of 30\%. Future experiments could show that an inclusion rate higher than 30\% is optimal but for the purpose of this experiment, it will suffice.

For the sake of simplicity, validation and test sets have been merged into one. For each dataset, there are 40000 images in the training set and 1000 in the test/validation split. This makes it an average sized data set which means that the inclusion rates on these datasets should reflect to other datasets of similar size with only minor deviation.

@@

We found that the loss dropped immediately to a reasonable value within the first five epochs but that the images produced were not of a high quality. The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile. Better results could have been achieved by training more models, each with a different inclusion of local data. This would allow us to draw a definitive correlation of local data to loss but in this case, the samples we have taken will give us an approximate idea of the optimal parameters.

@@

After all models have been trained on their respective datasets, we evaluate them on the purely South African dataset. This gives us a good understanding of how much local data to include with foreign datasets if we want reasonable results. The solely South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split. These were both used to construct the hybrid datasets. Now, we evaluate the model that has been previously trained on the foreign-local hybrid set on a purely South African dataset and measure its performance.

We expect the loss to be much larger when very few South African images have been included into the hybrid set. As we evaluate datasets with a larger inclusion of local data, we expect the models loss to decrease. Using this data, we can estimate the optimal inclusion of local data for satisfactory results when performing machine learning in foreign countries.

@@

Clearly, the difference in loss is minuscule. The small drop in loss that you get when including local data does not warrant the time and effort spent retrieving, processing and preparing at first. The majority of datasets used in image colourisation or classification already contain a wide variety of images that will make including local data redundant. By no means does this imply that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin. It does, however, mean that the inclusion of local data merely does not effect the outcome of the experiment as much as other factors would. Other factors include running the experiment for longer or simply finding a larger dataset. The insignificance of local data inclusion is made extremely evident when comparing images that were produced using a model trained without any local data with a model trained using 30\% local data,

@@

Many studies have shown that the majority of datasets used in facial analysis are comprised of mostly lighter skinned subjects. Research published by Joy Buolamwini \cite{buolamwini2018gender} shows that when evaluated on industry standard facial analysis networks, dark skinned females yield the highest error rates (of up to 34.7\%). This is alarming considering that light skinned males yield a maximum error rate of only 0.8\% when evaluated on the same dataset. \cite{corbett2018measure}

@@

No substantial research has been conducted in investigating the effect of data imbalance with respect to image colourisation. This may be due to the fact that image colourisation is largely subjective as made evident in the previous section.

@@

Evidently, the errors in the images coloured for the various percentage inclusions of local data are relatively the same - only a 3.3\% improvement was recorded for the training of the model from the 0\% inclusion to the 30\% inclusion. Such an improvement is mitigated when considering that the error on the colourisation of the images is already of an order of magnitude of $-3$, and is likely only to be noticed by computers and those utilising computational resources to analyse the images.

@@

Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect. Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is done solely by computers, with a corresponding report being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.

Therefore, the result of whether or not dataset bias exists is conclusive: it does and the inclusion of more diverse data in the set improves the performance of the model; the result of whether or not the building of a more diverse dataset is inconclusive: this would be decided on a per-application basis and what relative accuracy is required.

>>>>

Sufficient research exists for the implementation of bias-mitigating machine learning algorithms. However, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general. To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise greyscale images of a South African context. There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to seventy-five epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after five epochs. This did lead to the hues of the images being slightly more South African in essence, but it was largely undetectable to the human eye. Consequently, the inclusion of local data in the dataset definitely minimises the bias of the dataset, but whether or not that bias-mitigation justifies the vast resources required to build the comprehensive dataset is determined on a case-by-case basis.

@@

The identified cause is the lack of diversity in the training datasets, because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being initially used, particularly when a non-South African dataset is used to train a model to operate on a South African sample.

Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset. However, constructing and testing datasets to determine such a goal is both time-consuming and expensive in terms of financial and computational resources required. Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both international and South African settings. This provides a simple way to build a diverse dataset that does not need to be labelled, or require excessive computational time, since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.

@@

For simplicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space. This makes it convenient when separating the grey and colour channels, since the 'L' channel in LAB is simply the greyscale image. This greyscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image. These 'AB' channels make up the colour and simply signify the red-green and blue-yellow shift respectively. HSB would be another option when choosing a suitable colour space. In general, the choice of colour space makes no difference to the resulting image that our network produces.

@@

When colourising images using a deep learning approach, it is generally easiest to develop the model as a regression problem. Classification would yield better results, more life-like and saturated, whereas regression would lead to output images appearing dull and bland. This is the result of the multi-modality of pictures: there is no clear colour for any one image. This leads our model, using an aggregate loss function, to simply find an average colour, resulting in a bland image.

@@

For the sake of simplicity, we continue to use regression, as it leads to satisfactory results while being simple to understand. We use a mean squared error loss which may average colour distribution, but still produces images with a reasonable amount of colour. If we aimed to produce images that are saturated and more diverse in colour, regression would not suffice. Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the output and the representivity of local images in the data, which it does.

@@

Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al. \cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colourisation as a classification problem.

Our network topology consists of the first six layers of a ResNet-18, followed by three layers of deconvolutions. We find that this simple structure works well enough for our purposes and is relatively fast. Our model uses the Adam optimiser alongside PyTorch's autograd, which makes backpropogating in our network trivial.

@@

In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40~000 images and being constructed with a different percentage of South African images from 0\% to 30\% inclusive, increasing in graduations of 10\%. In addition to this training dataset, a validation dataset (of 1000 images in total) is also prepared with the same corresponding percentage inclusion of South African images.

Each version of the dataset will be used to train the model independently of each other for a total of seventy-five epochs each, with validation being done at the conclusion of each epoch.

A set collection of South African and international images will be passed through each model respectively, with the subsequent losses being an indication of the bias present in the model, due to the predominant locale present in the training images. (We take the dataset version with 0\% South African images included as the base case of how the model performs on South African image evaluation when trained exclusively on international images.)

@@

The collection of South African images\footnote{All images are freely available with a Creative Commons Licence or are in the public domain} was downloaded from the image-sharing service, Flickr. Only images that have been tagged with one of the following keywords were selected:

@@

A Python script was used to assimilate the international and South African images into the required ratios (of 0\% to 30\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split. The same number of images ($x$\%) was removed from the foreign dataset, meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30\}$

@@

For each dataset, we trained our convolutional model from the beginning with the same hyperparameters. The following table illustrates the hyperparameters used:

@@

Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a higher inclusion rate to the above mentioned epoch. For this reason, we chose to limit our inclusion of South African data to a maximum of 30\%. Future experiments could show that an inclusion rate higher than 30\% is optimal, but for the purpose of this experiment, it will suffice.

For the sake of simplicity, validation and test sets have been merged into one. For each dataset, there are 40~000 images in the training set and 1000 in the test/validation split. This makes it an average sized dataset, which means that the inclusion rates on these datasets should reflect to other datasets of similar size with only minor deviation.

@@

We found that the loss dropped immediately to a reasonable value within the first five epochs, but that the images produced were not of a high quality. The training process was long and slow with very little change in loss from one epoch to the next, but the end results make it worthwhile. Better results could have been achieved by training more models, each with a different inclusion of local data. This would allow us to draw a definitive correlation of local data to loss, but in this case, the samples we have taken will give us an approximate idea of the optimal parameters.

@@

After all models had been trained on their respective datasets, we evaluated them on the purely South African dataset. This gave us a good understanding of how much local data to include with foreign datasets if we wanted reasonable results. The solely South African dataset contains 35~000 images in its training split and 5000 images in the validation/testing split. These were both used to construct the hybrid datasets. Then we evaluated the model that had been previously trained on the foreign-local hybrid set on a purely South African dataset and measured its performance.

We expected the loss to be much larger when very few South African images had been included into the hybrid set. As we evaluate datasets with a larger inclusion of local data, we expect the models' loss to decrease. Using the data from the models that we trained, we can estimate the optimal inclusion of local data for satisfactory results when performing machine learning in foreign countries.

@@

Clearly, the difference in loss is minuscule. The small drop in loss that you get when including local data does not warrant the time and effort spent retrieving, processing and preparing at first. The majority of datasets used in image colourisation or classification already contain a wide variety of images that will make including local data redundant. By no means does this imply that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin. It does, however, mean that the inclusion of local data merely does not affect the outcome of the experiment as much as other factors would. Other factors include running the experiment for longer or simply finding a larger dataset. The insignificance of local data inclusion is made extremely evident when comparing images that were produced using a model trained without any local data with a model trained using 30\% local data,

@@

Many studies have shown that the majority of datasets used in facial analysis comprise mostly lighter-skinned subjects. Research published by Joy Buolamwini \cite{buolamwini2018gender} shows that when evaluated on industry standard facial analysis networks, dark-skinned females yield the highest error rates (of up to 34.7\%). This is alarming considering that light-skinned males yield a maximum error rate of only 0.8\% when evaluated on the same dataset. \cite{corbett2018measure}

@@

No substantial research has been conducted in investigating the effect of data imbalance with respect to image colourisation. This may be due to the fact that image colourisation is largely subjective, as made evident in the previous section.

@@

Evidently, the errors in the images coloured for the various percentage inclusions of local data are relatively the same - only a 3.3\% improvement was recorded for the training of the model from the 0\% inclusion to the 30\% inclusion. Such an improvement is mitigated when considering that the error on the colourisation of the images is already of an order of magnitude of $-3$, and is likely only to be noticed by computers and those using computational resources to analyse the images.

@@

Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect. There are inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is done solely by computers, with a corresponding report being generated. In such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.

Therefore, the result of whether or not dataset bias exists is conclusive: it does and the inclusion of more diverse data in the set improves the performance of the model. The need to build a more diverse dataset should be decided on a per-application basis depending on what relative accuracy is required.
